Stage-Splitting a Modal Language

### 2014-07-31 paper/semantics.tex
added last step in rowan's semantics example
be clear about which language residuals and values are in

If we simply treat \verb|(next {1+2}, 7)| as a value during stage~\bbone, and
substitute it for the three occurrences of \verb|x| in the body of the
\verb|let| expression, the result of stage \bbone\ computation is
\begin{lstlisting} 
prev{
  next{ prev{next {1+2}} * prev{next {1+2}} * 7 }
}
\end{lstlisting}
Stage-\bbtwo\ evaluation of this expression will compute \verb|1+2| twice. Of
course, this is not particularly expensive, but in general, the duplicated
computation could be anything, even a recursive call to the function! To avoid duplicating
computations, we take a different approach.  Instead of
duplicating the contents of the $\next$ expression in the tuple, we bind the contained stage-\bbtwo\ expression to
a variable (here, $\mathtt{y}$) and duplicate only a reference to that variable.
This produces:
\begin{lstlisting} 
let y = 1+2 in y * y * 7
\end{lstlisting}

@@

The algorithm described above creates three different kinds of expressions which
cannot be evaluated further at a particular stage:
\begin{itemize}
\item 
Partial values ($\pvalsym$s) are stage \bbone\ terms that have been fully evaluated, 
but which may contain stage-\bbtwo\ variables wrapped in $\next$ blocks. 
In the example above, 
\verb|(|$\next \{\mathtt{\hat y}\}$\verb|,7)|.
\item Residuals ($\ressym$es) are stage \bbtwo\ terms whose stage \bbone\
subexpressions have all been fully evaluated. In the example above,
\verb|(1+2)| and \verb|(let z = y*y*7 in z)|.
\item Values ($\valsym$s) are stage \bbtwo\ terms which are fully evaluated; these
are the results of a computation after both stages have been completed. In the
example above, the term evaluates to \verb|63|.
\end{itemize}
The $\redonesym$ judgment takes an open stage \bbone\ term to a {\em future environment} $\xi$ and a partial value $v$.
The future environment is a mapping from newly-created variables 
(which then may appear within $\next$ blocks in $v$)
to residuals.  It is the formal manifestation of the floating substitutions in our example.
For all of the normal features of \lang (\ref{fig:diaSemanticsCore}), first stage evaluation has the same behavior and effect on values as standard evaluation,
and the final future environment is gotten by merging the future environments of the subterms.

>>>>

If we simply treat \verb|(next {1+2}, 7)| as a value during stage~\bbone, and
substitute it for the three occurrences of \verb|x| in the body of the
\verb|let| expression, the result of stage \bbone\ computation is
\begin{lstlisting} 
prev{
  next{ prev{next {1+2}} * prev{next {1+2}} * 7 }
}
\end{lstlisting}
Finally, the $\prev$s eliminate the $\next$s to yield the final residual:
\begin{lstlisting} 
(1+2) * (1+2) * 7
\end{lstlisting}
Stage-\bbtwo\ evaluation of this expression will compute \verb|1+2| twice. Of
course, this is not particularly expensive, but in general, the duplicated
computation could be anything, even a recursive call to the function! To avoid duplicating
computations, we take a different approach.  Instead of
duplicating the contents of the $\next$ expression in the tuple, we bind the contained stage-\bbtwo\ expression to
a variable (here, $\mathtt{y}$) and duplicate only a reference to that variable.
This produces:
\begin{lstlisting} 
let y = 1+2 in y * y * 7
\end{lstlisting}

@@

To fully evaluate an expression in \lang, we first evaluate all of its stage
\bbone\ subexpressions as described above. This results in a stage \bbtwo\
expression with no stage \bbone\ content; we say that it is a term in a
monostaged%
\footnote{In general, \lang could have more than two stages, in which case the
expression would be in a language with one fewer stage.}
language called \langTwo. Ordinary \langTwo\ evaluation, implemented by the
$\tworedsym$ judgment, completes the evaluation of this term.

The algorithm described above creates three different kinds of expressions which
cannot be evaluated further at a particular stage:
\begin{itemize}
\item 
Partial values ($\pvalsym$s) are stage \bbone\ terms that have been fully evaluated, 
but which may contain stage-\bbtwo\ variables wrapped in $\next$ blocks. 
In the example above, 
\verb|(|$\next \{\mathtt{\hat y}\}$\verb|,7)|.
\item Residuals ($\ressym$es) are \langTwo\ terms---stage \bbtwo\ terms whose
stage \bbone\ subexpressions have all been fully evaluated. In the example
above,
\verb|(1+2)| and \verb|(let z = y*y*7 in z)|.
\item Values ($\valsym$s) are \langTwo\ terms which are fully evaluated; these
are the results of a computation after both stages have been completed. In the
example above, the term evaluates to \verb|63|.
\end{itemize}
The $\redonesym$ judgment takes an open stage \bbone\ term to a {\em future
environment} $\xi$ and a partial value $v$.  The future environment is a mapping
from fresh stage \bbtwo\ variables (which may appear inside $\next$ blocks in
$v$) to residuals---in our example, we represented this environment with
explicit substitutions. For all of the normal features of \lang
(\ref{fig:diaSemanticsCore}), first stage evaluation has the same behavior and
effect on values as standard evaluation, and the final future environment is
gotten by merging the future environments of the subterms.
